Learning and Evaluating \strips\ Action Models with Classical Planning

### 2018-03-13 doc/aij18/aij18.tex
updates added to the aij18 paper

\title{Learning and Evaluation of \strips\ Action Models with Classical Planning}
\author[label1]{Diego Aineto}
\author[label1]{Sergio Jim\'{e}nez Celorrio}
\author[label1]{Eva Onaindia}
\address[label1]{Department of Computer Systems and Computation, Universitat Polit\`ecnica de Val\`encia. Spain}

@@

\begin{abstract}
  This paper presents a novel approach for learning \strips\ action models from observations of plan executions that compiles this learning task into classical planning. The compilation approach is flexible to various amount and kind of available input knowledge; the learning examples can range from plans (with their corresponding initial and final states), sequences of state observations or just a set of initial and final states (where no intermediate action or state is given). The compilation accepts also partially specified action models and can be used to validate whether the observation of a plan execution follows a given \strips\ action model, even if this model is not fully specified. What is more, the compilation is extensible to assess how well a \strips\ action model matches observations of plan executions. The performance of our compilation approach is evaluated learning action models for a wide range of classical planning domains from the International Planning Competition (IPC) and assessing the learned models with respect to (1), test sets of observations of plan executions and (2), the true models.
\end{abstract}

@@

Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models, the vanilla action model for planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The {\em input} to ML algorithms (the learning/training data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each plan possibly has a different length and involve a different number of objects).
\item The {\em output} of ML algorithms usually is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of preconditions, negative and positive effects, that define the possible state transitions.
\end{itemize}

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016generalized,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. A solution to the classical planning task that results from our compilation is a sequence of actions that determines the learned action model, i.e. the preconditions and effects of the target \strips\ operator schemas.

The compilation approach is appealing by itself, because it leverages off-the-shelf planners and because it opens up a way towards \emph{bootstrapping} planning action models, enabling a planner to gradually learn/update its action model. Moreover, the practicality of the compilation allow us to report learning results over a wide range of IPC planning domains. Apart from these, the compilation approach presents the following contributions:
\begin{enumerate}
\item {\em Flexibility in the input knowledge}. The compilation is flexible to various amounts and kind of available input knowledge. Learning examples can range from a set of plans (with their corresponding initial and final states) or state observations, to just a set of initial and final states where no intermediate action or state is observed. The compilation also accepts previous knowledge about the structure of the actions in the form of partially specified action models, where some preconditions and effects are a priori known.

\item {\em Model Evaluation}. The compilation can assess how well a \strips\ action model matches a given set of observations of plan executions. Our compilation is extensible to accept a learned model as input besides the observations of plan executions. This extension allows us to transform the input model into a new model that induces the observations whilst assessing the amount of edition required by the input model to induce the given observations. 
\end{enumerate}

A first description of the compilation previously appeared in several conference papers~\cite{aineto2018learning}. Compared to the conference papers, the present paper includes the following novel material:
\begin{itemize}
\item Unifies the formulation of our approach for learning \strips\ action models from state observations and from observed plans.
\item Shows that our compilation is also valid for the particular scenario where the observed states are not full states but {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing.
\item Proposes a distance metric to assess how well a \strips\ action model matches a given set of plans. Defines the compilation to compute this distance as well as the edit distance between two given \strips\ action models.
\end{itemize}

Section~\ref{sec:Section2} reviews related work on learning planning action models. Section~\ref{sec:Section3} formalizes the classical planning model with {\em conditional effects} (a requirement of the proposed compilation) and the \strips\ action model (the output of the addressed learning task). Section~\ref{sec:Section4} formalizes the learning of \strips\ action models with regard to different amounts of available input knowledge. Sections~\ref{sec:Section5} and ~\ref{sec:Section6} describe our compilation approach for addressing the formalized learning tasks and how the compilation is also extensible to assess learned action models. Section~\ref{sec:Section7} reports the data collected in a two-fold empirical evaluation of our learning approach: First, the learned \strips\ action models are tested with observations of plan executions and second, the learned models are compared to the actual models. Finally, Section ~\ref{sec:Section8} discusses the strengths and weaknesses of the compilation approach and proposes several opportunities for future research.

@@

@@

A {\em classical planning frame} is a tuple $\Phi=\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions. An action $a\in A$ is defined with {\em preconditions}, $\pre(a)\subseteq\mathcal{L}(F)$, {\em positive effects}, $\eff^+(a)\subseteq\mathcal{L}(F)$, and {\em negative effects} $\eff^-(a)\subseteq\mathcal{L}(F)$. We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

@@

We assume also that actions $a\in A$ are instantiated from \strips\ operator schemas $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ where:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, is the operator {\em header} defined by its name and the corresponding {\em variable names}, $pars(\xi)=\{v_i\}_{i=1}^{ar(\xi)}$. The headers of a four-operator {\em blocksworld} are {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the negative effects $del(\xi)\subseteq F_v$, and the positive effects $add(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}
Therefore, given the set of predicates $\Psi$ and the header of the operator schema $\xi$, then $2^{3|F_v(\xi)|}$ is an upper-bound on the number of possible \strips\ action model for $\xi$. In practice this number is lower and is given by $2^{2|F_v(\xi)|}$ because of the previous \strips\ constraints that require that negative effects appear as preconditions and that they cannot be positive effects and also, that a positive effect cannot appear as a precondition.

Last but not least, we say that two \strips\ action models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff (1) the operator schemes of both models have the same headers and (2), both are built from the same set of possible elements $F_v$. In other words, for every $\xi\in\mathcal{M}$ and $\xi'\in\mathcal{M'}$ such that $head(\xi)=head(\xi')$, it holds that $F_v(\xi)=F_v(\xi')$.

@@

\subsection{Learning from observations of plan executions}

The first learning task corresponds to observing an agent acting in the world but watching only the states that result of its plan executions, the actual executed actions are unobserved. This learning task is formalized as $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$:
\begin{itemize}
\item $\mathcal{M}$ is the set of {\em empty} operator schemas, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$. In some cases, we may not require to start learning from scratch that is, the operator schemas in $\mathcal{M}$ may be not {\em empty} but {\em partially specified} operator schemes where some preconditions and some positive or negative effects are a priori known.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The states in $\mathcal{O}$ are not required to be full states, they can be {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing (it is unknown whether their value is either positive or negative). We assume that observations are noiseless, meaning that if the value of a fluent is observed it is correct.
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning frame. This set of predicates can be given as input or inferred from the state observations given in $\mathcal{O}$ provided that they are full states.
\end{itemize}

A solution to the $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ learning task is a set of operator schema $\mathcal{M}'$ compliant with the headers in $\mathcal{M}$, the state observation sequence $\mathcal{O}$ and the predicates $\Psi$. In this learning scenario, a solution must not only determine a possible \strips\ action model but also the plan $\pi$, that explains the given observations using the learned \strips\ model. Figure~\ref{fig:example-observations} shows a $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ task for learning the {\em blocksworld} \strips\ action model from the five-state observations sequence that corresponds to inverting a 2-block tower.

@@

Here we redefine the $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ learning task to cover the scenario where the actions executed by the observed agent are known. Here $\mathcal{O}$ may contain partial states because otherwise, the learning task becomes trivial as explained. In this case, the learning task is formalized as $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, where:
\begin{itemize}
\item The plan $\pi=\tup{a_1, \ldots, a_n}$, is an action sequence that produces the sequence of state observations given in $\mathcal{O}$. Note that if the plan is {\em diverse} enough, i.e. it contains at least one ground action for each of the aimed action schemes, it can be used to infer the set of {\em empty} operator schemas $\mathcal{M}$.
\end{itemize}

Figure~\ref{fig:example-plans} shows an example of a {\em blocksworld} learning task $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, that corresponds to observing the execution of an four-action plan for inverting a two-block tower. Note that $\mathcal{M},\Psi$ are skipped since they are the same as for Figure~\ref{fig:example-observations}.

@@

The previous definitions formalize the learning \strips\ action models from observations of a single plan execution. These definitions are extensible to the more general case where learning from the execution of multiple plans:
\begin{itemize}
  \item When learning from states observations, $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ is still a valid formalization by simply considering that now $\mathcal{O}=\tup{s_0^1,s_1^1,\ldots,s_{n}^1,\ldots,s_0^t,s_1^t,\ldots,s_{n}^t\ldots,s_0^{\tau},s_1^{\tau},\ldots,s_{n}^{\tau}}$ are sequences of {\em state observations} obtained observing the execution of multiple {\em unobserved} plans $\pi^t=\tup{a_1, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$} such that, for each {\small $1\leq i\leq n^t$}, $a_i^t$ is applicable in $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
  \item When learning from a set of observed plans, the task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\Pi}$, where $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is the given set of example plans producing the corresponding set of state observations $\mathcal{O}$.
\end{itemize}

\section{Learning \strips\ action models with classical planning}
\label{sec:Section5}
Our approach for addressing a $\Lambda$ learning task is compiling it into a classical planning task $P_{\Lambda}$ with conditional effects. A planning compilation is a suitable approach because computing a solution for $\Lambda$ involves, not only determining the \strips\ action model $\mathcal{M}'$, but also the {\em unobserved} plans that explain the inputs to the learning task. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the action model $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\mathcal{M}$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the action model $\mathcal{M}'$}. The solution plan continues with a postfix that reproduces the given input knowledge (the available observations of the plan executions) using the programmed action model $\mathcal{M}'$.
\end{enumerate}

\subsection{Learning from state observations}
Here we formalize the compilation for learning \strips\ action models with classical planning. Given a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects $\Omega$ appearing in the input observations $\mathcal{O}$, i.e. {\tt\small block A} and {\tt\small block B} in Figure~\ref{fig:example-observations}. Formally, $\Omega=\bigcup_{\small s\in\mathcal{O}} obj(s)$, where $obj$ is a function that returns the objects that appear in a given state.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in the schema $\xi\in \mathcal{M}'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item The fluents $mode_{prog}$ and $mode_{val}$ indicating whether the operator schemas are programmed or validated and the fluents $\{test_i\}_{1\leq i\leq |\mathcal{O}|}$, indicating the observation in $\mathcal{O}$ where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. Our compilation assumes that initially, operator schemas are programmed with every possible precondition (the most specific learning hypothesis), no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.

@@

\item Actions for {\em applying} a programmed operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Given that the operators headers are known, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}

>>>>

\title{Learning and Evaluating \strips\ Action Models with Classical Planning}
\author[label1]{Diego Aineto}
\author[label1]{Sergio Jim\'{e}nez Celorrio}
\author[label1]{Eva Onaindia}
\address[label1]{Department of Computer Systems and Computation, Universitat Polit\`ecnica de Val\`encia. Spain}

@@

\begin{abstract}
  This paper presents a novel approach for learning \strips\ action models from observations of plan executions that compiles this learning task into classical planning. The compilation approach is flexible to various amount and kind of available input knowledge; learning examples can range from plans (with their corresponding initial state), sequences of state observations or just a set of initial and final states (where no intermediate action or state is a priori known). The compilation accepts also partially specified action models and can be used to validate whether the observation of a plan execution follows a given \strips\ action model, even if this model is not fully specified. What is more, the compilation is extensible to assess how well a given \strips\ action model matches a {\em test set} with observations of plan executions. The performance of our compilation approach is evaluated learning action models for a wide range of classical planning domains from the International Planning Competition (IPC) and assessing the learned models with respect to (1), test sets of observations of plan executions and (2), the true models.
\end{abstract}

@@

Machine Learning (ML) has shown to be able to compute a wide range of different kinds of models from examples~\cite{michalski2013machine}. The application of inductive ML to the learning of \strips\ action models, the vanilla action model for planning~\cite{fikes1971strips}, is not straightforward though:
\begin{itemize}
\item The {\em input} to ML algorithms (the {\em learning/training} data) usually are finite vectors encoding the value of fixed features in a given set of objects. The input for learning planning action models are observations of plan executions (where each plan possibly has a different length and may involve a different number of objects).
\item The {\em output} of ML algorithms usually is a scalar value (an integer, in the case of {\em classification} tasks, or a real value, in the case of {\em regression} tasks). When learning \strips\ action models the output is, for each action, the sets of {\em preconditions}, {\em negative} and {\em positive effects} that define the possible state transitions.
\end{itemize}

Motivated by recent advances on the synthesis of different kinds of generative models with classical planning~\cite{bonet2009automatic,segovia2016generalized,segovia2016hierarchical,segovia2017generating}, this paper introduces an innovative approach for learning \strips\ action models that can be defined as a classical planning compilation. A solution to the classical planning task that results from our compilation is a sequence of actions that determines the learned action model, i.e. the preconditions and effects of the target \strips\ actions.

The compilation approach is appealing by itself because it leverages off-the-shelf planners for learning and because its practicality allow us to report model learning results over a wide range of IPC planning domains. Moreover, it opens up a way towards \emph{bootstrapping} planning action models, enabling a planner to gradually learn/update its action model. Apart from these, the compilation approach presents the following contributions:
\begin{enumerate}
\item {\em Flexibility}. The compilation is flexible to various amount and kind of available input knowledge. Learning examples can range from a set of plans (with their corresponding initial state) or state observations, to just a set of initial and final states where no intermediate action or state is observed. Last but not least, the compilation also accepts previous knowledge about the structure of the actions in the form of partially specified action models, where some preconditions and effects are a priori known. % Also accepts background knowledge

\item {\em Model Evaluation}. The compilation can assess how well a given \strips\ action model matches a {\em test set} with observations of plan executions. Our compilation is extensible to accept a learned model as input besides the observations of plan executions. This extension allows us to transform the input model into a new model that induces the observations whilst assessing the amount of edition required by the input model to induce the given observations. 
\end{enumerate}

A first description of the compilation previously appeared in the conference paper~\cite{aineto2018learning}. Compared to the conference paper, this work includes the following novel material:
\begin{itemize}
\item Unifies the formulation of our approach for learning \strips\ action models from observed plans but also from state observations.
\item Evaluates our compilation in the particular scenario where the observed states are not full states but {\em partial states}. This mean that some of the observed fluents (either with {\em positive} or {\em negative} value) are missing because they cannot be observed.
\item Proposes a distance metric to assess how well a \strips\ action model matches a given set of plans. Extends the compilation to compute this distance as well as the edit distance between two given \strips\ action models. In addition defines the task of {\em model recognition} as the task of selecting from a given set, the action model that best matches a {\em test set} with observations of plan executions.
\item Shows that learning \strips\ action models can also leverage {\em background knowledge} in this case, given as planning constraints, either in the form of {\em state constraints} or {\em trajectory constraints}.
\end{itemize}

Section~\ref{sec:Section2} reviews related work on learning planning action models. Section~\ref{sec:Section3} introduces the classical planning model with {\em conditional effects} (a requirement of the proposed compilation) and the \strips\ action model (the output of the addressed learning task). Section~\ref{sec:Section4} formalizes the learning of \strips\ action models with regard to different amount and kind of available input knowledge. Sections~\ref{sec:Section5} and ~\ref{sec:Section6} describe our compilation approach for addressing the formalized learning tasks and how the compilation is also extensible to assess learned action models. Section~\ref{sec:Section7} reports the data collected in a two-fold empirical evaluation of our learning approach: First, the learned \strips\ action models are tested with observations of plan executions and second, the learned models are compared to the actual models. Finally, Section ~\ref{sec:Section8} discusses the strengths and weaknesses of the compilation approach and proposes several opportunities for future research.

@@

Recenctly classical planning compilations have been defined to learn different kinds of generative models from examples. The existing compilations for computing FSCs for generalized planning follow a {\it top-down} approach that interleaves {\it programming} the FSC with validating it and hence, they tightly integrate planning and generalization. To keep the computation of FSCs tractable, they limit the space of possible solutions bounding the maximum size of the FSC. In addition, they impose that the instances to solve share, not only the domain theory (actions and predicates schemes) but the set of fluents~\cite{javi-Gplanning-ICAPS16} or a subset of {\it observable} fluents~\cite{Geffner:FSM:AAAI10}. Programs increase the readability of FSCs separating the control-flow structures from the primitive actions. Like FSCs, programs can also be computed following a {\it top-down} approach, e.g.~exploiting compilations that program and validate the program on instances with the same state and action space~\cite{javi-Gplanning-ICAPS16}. Since these {\it top-down} approaches search in the space of solutions, it is helpful to limit the set of different control-flow instructions. For instance using only {\it conditional gotos} that can both implement branching and loops~\cite{Jimenez15}.

@@

A {\em classical planning frame} is a tuple $\Phi=\tup{F,A}$, where $F$ is a set of fluents and $A$ is a set of actions. An action $a\in A$ is defined with {\em preconditions}, $\pre(a)\subseteq\mathcal{L}(F)$, {\em negative effects} $\eff^-(a)\subseteq\mathcal{L}(F)$ and {\em positive effects}, $\eff^+(a)\subseteq\mathcal{L}(F)$. We say that an action $a\in A$ is {\em applicable} in a state $s$ iff $\pre(a)\subseteq s$. The result of applying $a$ in $s$ is the {\em successor state} denoted by $\theta(s,a)=\{s\setminus\eff^-(a))\cup\eff^+(a)\}$.

@@

We assume also that actions $a\in A$ are instantiated from \strips\ operator schemas $\xi=\tup{head(\xi),pre(\xi),add(\xi),del(\xi)}$ where:
\begin{itemize}
\item $head(\xi)=\tup{name(\xi),pars(\xi)}$, is the operator {\em header} defined by its name and the corresponding {\em variable names}, $pars(\xi)=\{v_i\}_{i=1}^{ar(\xi)}$. The headers of a four-operator {\em blocksworld} are {\small\tt pickup($v_1$), putdown($v_1$), stack($v_1,v_2$)} and {\small\tt unstack($v_1,v_2$)}.
\item The preconditions $pre(\xi)\subseteq F_v$, the negative effects $del(\xi)\subseteq F_v$, and the positive effects $add(\xi)\subseteq F_v$ such that, $del(\xi)\subseteq pre(\xi)$, $del(\xi)\cap add(\xi)=\emptyset$ and $pre(\xi)\cap add(\xi)=\emptyset$.
\end{itemize}
Therefore, given the set of predicates $\Psi$ and the header of the operator schema $\xi$, then $2^{3|F_v(\xi)|}$ is an upper-bound on the number of possible \strips\ action models for $\xi$. In practice this upper-bound is lower, $2^{2|F_v(\xi)|}$, given that the previous \strips\ constraints require that negative effects appear as preconditions and that they cannot be positive effects and also, that a positive effect cannot appear as a precondition.

Last but not least, we say that two \strips\ operator schemes $\xi$ and $\xi'$ are {\em comparable} if both schemas have the same headers and are built from the same set of possible elements. Formally, iff $head(\xi)=head(\xi')$ and it holds that $F_v(\xi)=F_v(\xi')$. Likewise we can say that two \strips\ action models $\mathcal{M}$ and $\mathcal{M}'$ are {\em comparable} iff, every $\xi\in\mathcal{M}$ is comparable with an action schema $\xi'\in\mathcal{M'}$ and viceversa, every $\xi'\in\mathcal{M'}$ is comparable with an action schema $\xi\in\mathcal{M}$.

@@

\subsection{Learning from observations of plan executions}
The first learning task corresponds to observing an agent acting in the world but watching only the states that result of its plan executions, the actual executed actions are unobserved. This learning task is formalized as $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$:
\begin{itemize}
\item $\mathcal{M}$ is the set of {\em empty} operator schemas, wherein each $\xi\in\mathcal{M}$ is only composed of $head(\xi)$. In some cases, we may not require to start learning from scratch that is, the operator schemas in $\mathcal{M}$ may be not {\em empty} but {\em partially specified} operator schemes. This means that some preconditions and some positive or negative effects are a priori known.
\item $\mathcal{O}=\tup{s_0,s_1,\ldots,s_{n}}$ is a sequence of {\em state observations} obtained observing the execution of an {\em unobserved} plan $\pi=\tup{a_1, \ldots, a_n}$ such that, for each {\small $1\leq i\leq n$}, $a_i$ is applicable in $s_{i-1}$ and generates the successor state $s_i=\theta(s_{i-1},a_i)$. The states in $\mathcal{O}$ are not required to be full states, they can be {\em partial states}, in the sense that some of the observed fluents (either with positive or negative value) are missing (it is unknown whether their value is either positive or negative). Anyway we assume that observations are noiseless, meaning that if the value of a fluent is observed it is correct.
\item $\Psi$ is the set of predicates that define the abstract state space of a given planning frame. This set of predicates can be given as input but it can also be inferred from the state observations given in $\mathcal{O}$ provided that they are full states.
\end{itemize}

A solution to the $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ learning task is a set of operator schema $\mathcal{M}'$ compliant with the input model $\mathcal{M}$, the state observation sequence $\mathcal{O}$ and the predicates $\Psi$. In this learning scenario, a solution must not only determine a possible \strips\ action model but also the unobserved plan $\pi$, that explains the given state observations using the learned \strips\ model. Figure~\ref{fig:example-observations} shows a $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ task for learning the {\em blocksworld} \strips\ action model from the five-state observations sequence that corresponds to inverting a 2-block tower. In this case we are assuming full state observability so, if a fluent is not present in the observations, it means that it is set to {\em false}.

@@

Here we redefine the learning task to cover the scenario where the actions executed by the observed agent are known. In this case $\mathcal{O}$ should contain partial states because otherwise, the learning task becomes trivial as explained. The learning task is now formalized as $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, where:
\begin{itemize}
\item The plan $\pi=\tup{a_1, \ldots, a_n}$, is an action sequence that produces the sequence of state observations given in $\mathcal{O}$. When the plan is {\em diverse} enough, i.e. it contains at least one ground action for each of the aimed action schemes, the set of {\em empty} operator schemas $\mathcal{M}$ can be inferred from the input plan $\pi$.
\end{itemize}

Figure~\ref{fig:example-plans} shows an example of a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi,\pi}$, that corresponds to observing the execution of an four-action plan for inverting a two-block tower. Note that $\mathcal{M},\Psi$ are skipped since they are the same as for Figure~\ref{fig:example-observations}. Again, we are assuming full state observability and the plan is full observable as well. However, only the first and last states were observed.

@@

The previous definitions formalize the learning of \strips\ action models from the observation of a single plan execution. These definitions are extensible to the more general case where the available input data is obtained from the execution of multiple plans:
\begin{itemize}
\item When learning from states observations, $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ is still a valid formalization by simply considering that now $\mathcal{O}=\tup{s_0^1,s_1^1,\ldots,s_{n}^1,\ldots,s_0^t,s_1^t,\ldots,s_{n}^t\ldots,s_0^{\tau},s_1^{\tau},\ldots,s_{n}^{\tau}}$ is a sequences of {\em state observations} that are obtained observing the execution of multiple {\em unobserved} plans $\pi^t=\tup{a_1, \ldots, a_n^t}$, {\tt\small $1\leq t\leq \tau$}, one after the other. This means that, for each {\small $1\leq i\leq n^t$}, $a_i^t\in \pi^t$ is applicable in $s_{i-1}^t$ and generates the successor state $s_i^t=\theta(s_{i-1}^t,a_i^t)$.
  \item When learning from a set of observed plans, the task is defined as $\Lambda=\tup{\mathcal{M},\Psi,\mathcal{O},\Pi}$, where $\Pi=\{\pi_1,\ldots,\pi_{\tau}\}$ is the given sequence of example plans producing the corresponding sequence of state observations $\mathcal{O}$.
\end{itemize}

\section{Learning \strips\ action models with classical planning}
\label{sec:Section5}
Our approach for addressing a $\Lambda$ learning task is compiling it into a classical planning task $P_{\Lambda}$ with conditional effects. A planning compilation is a suitable approach because computing a solution for $\Lambda$ involves, not only determining the \strips\ action model $\mathcal{M}'$, but also the {\em unobserved} plans that explain the given inputs to the learning task. The intuition behind the compilation is that a solution to the resulting classical planning task is a sequence of actions that:

\begin{enumerate}
\item {\bf Programs the action model $\mathcal{M}'$}. A solution plan starts with a {\em prefix} that, for each $\xi\in\mathcal{M}$, determines which fluents $f\in F_v(\xi)$ belong to its $pre(\xi)$, $del(\xi)$ and $add(\xi)$ sets.
\item {\bf Validates the action model $\mathcal{M}'$}. The solution plan continues with a postfix that reproduces the given input knowledge (the available observations of the plan executions) but using the programmed action model $\mathcal{M}'$.
\end{enumerate}

\subsection{Learning from state observations}
Here we formalize the compilation for learning \strips\ action models with classical planning. Given a learning task $\Lambda=\tup{\mathcal{M},\mathcal{O},\Psi}$ the compilation outputs a classical planning task $P_{\Lambda}=\tup{F_{\Lambda},A_{\Lambda},I_{\Lambda},G_{\Lambda}}$:
\begin{itemize}
\item $F_{\Lambda}$ contains:
\begin{itemize}
\item The set of fluents $F$ built instantiating the predicates $\Psi$ with the objects $\Omega$ that appear in the input observations, i.e. {\tt\small block A} and {\tt\small block B} in Figure~\ref{fig:example-observations}. Formally, $\Omega=\bigcup_{\small s\in\mathcal{O}} obj(s)$, where $obj$ is a function that returns the objects that appear in a given state.
\item Fluents $pre_f(\xi)$, $del_f(\xi)$ and $add_f(\xi)$, for every $f\in F_v(\xi)$, that represent the programmed action model. If a fluent $pre_f(\xi)/del_f(\xi)/add_f(\xi)$ holds, it means that $f$ is a precondition/negative/positive effect in the schema $\xi\in \mathcal{M}'$. For instance, the preconditions of the $stack$ schema (Figure~\ref{fig:stack}) are represented by the pair of fluents {\small\tt pre\_holding\_stack\_$v_1$} and {\small\tt pre\_clear\_stack\_$v_2$} set to {\em True}.
\item The fluents $mode_{prog}$ and $mode_{val}$ to indicate whether the operator schemas are programmed or validated, and the fluents $\{test_i\}_{1\leq i\leq |\mathcal{O}|}$, indicating the observation in $\mathcal{O}$ where the action model is validated.
\end{itemize}
\item $I_{\Lambda}$ contains the fluents from $F$ that encode $s_0$ (the first observation) and $mode_{prog}$ set to true. Our compilation assumes that initially, operator schemas are programmed with every possible precondition (the most specific learning hypothesis), no negative effect and no positive effect. Therefore fluents $pre_f(\xi)$, for every $f\in F_v(\xi)$, hold also at the initial state.

@@

\item Actions for {\em applying} a programmed operator schema $\xi\in\mathcal{M}$ bound with objects $\omega\subseteq\Omega^{ar(\xi)}$. Since operators headers are given as input, the variables $pars(\xi)$ are bound to the objects in $\omega$ that appear at the same position. Figure~\ref{fig:compilation} shows the PDDL encoding of the action for applying a programmed operator $stack$ from {\em blocksworld}.
\begin{small}
\begin{align*}
\hspace*{7pt}\pre(\mathsf{apply_{\xi,\omega}})=&\{pre_{f}(\xi)\implies p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))}\cup \{\neg mode_{val}\},\\
\cond(\mathsf{apply_{\xi,\omega}})=&\{del_{f}(\xi)\}\rhd\{\neg p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{add_{f}(\xi)\}\rhd\{p(\omega)\}_{\forall p\in\Psi,f=p(pars(\xi))},\\
&\{mode_{prog}\}\rhd\{\neg mode_{prog}\},\\
&\{\emptyset\}\rhd\{mode_{val}\}.
\end{align*}
\end{small}
